---
title: "Condition-aware proteomics missingness with ImputeFinder"
author:
- name: Emir Turkes
  email: emir.turkes@eturkes.com
package: imputefinder
output:
  BiocStyle::html_document:
    toc_float: true
abstract: |
  ImputeFinder classifies quantitative-proteomics missingness independently
  within experimental conditions. This vignette demonstrates the on/off rescue,
  manual and automatic intensity cutoffs, auditable result data, and the split
  into condition-specific downstream imputation branches.
vignette: |
  %\VignetteIndexEntry{Condition-aware proteomics missingness with ImputeFinder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4.5,
    dpi = 96,
    crop = NULL
)
library(imputefinder)
```

# Introduction

Missing quantitative-proteomics observations can reflect more than one
process. Some values are absent for approximately intensity-independent,
MAR-like reasons. Low-abundance features can instead fall below detection and
produce MNAR-like missingness. A protein can also be detected throughout one
biological condition and absent throughout another. That condition-specific
on/off pattern can be informative, but a single global label obscures it.

ImputeFinder addresses this boundary of the workflow. It takes unnormalised
log2 protein intensities, assigns a state to every feature-condition block,
filters features without adequate evidence, and returns the data and audit
trail needed for downstream processing. The labels mean **likely MAR** and
**likely MNAR** under an intensity-cutoff heuristic; they do not establish the
true causal missingness mechanism. ImputeFinder neither normalises nor imputes
values.

Bioconductor packages such as
[`DEP`](https://bioconductor.org/packages/DEP),
[`MSnbase`](https://bioconductor.org/packages/MSnbase), and
[`msImpute`](https://bioconductor.org/packages/msImpute) provide broader
proteomics analysis or imputation machinery. ImputeFinder has a narrower role:
it preserves condition-specific state and coordinates separate downstream
branches while returning either an ordinary matrix or a
`SummarizedExperiment`.

# Installation

Once ImputeFinder is available from a Bioconductor package repository, install
and load it as follows. Installation code is intentionally not evaluated while
building the vignette.

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("imputefinder")

library(imputefinder)
```

# Input and classification rules

The input must be an ordinary numeric matrix, or a selected numeric assay in a
`SummarizedExperiment`, with named protein features in rows and uniquely named
samples in columns. Missing values are `NA`; `NaN` and infinite values are
rejected. Condition labels must align one-to-one with samples. ImputeFinder
trusts the scientific input contract: it does not infer or apply a log
transform, normalisation, centring, or scaling, and it does not reinterpret
zero.

After rescue, each feature-condition block receives exactly one state:

| State | Evidence |
|---|---|
| `complete` | No value is missing. |
| `MNAR` | At least one value is missing and the arithmetic mean of observed values is below the condition cutoff. |
| `MAR` | At least one value is missing, the mean is at or above the cutoff, and strictly more than half the samples are observed. |
| `insufficient` | The block is on the MAR side of the cutoff but lacks a strict observed majority. |

The arithmetic mean is deliberate: one strong observed signal can move a
partly observed block toward the MAR side. Equality belongs to that side too.
MAR-like downstream estimation depends on relationships among observed values,
so a strict majority supplies minimum within-condition evidence. The majority
rule therefore applies only to MAR candidates. A sparse low-intensity block can
remain eligible as MNAR because one observed value may be the available
detection-limit evidence.

A feature is dropped if it is absent globally, `insufficient` in any condition,
or `MNAR` in every condition. Otherwise its complete state matrix is retained,
including mixtures such as MNAR in A and complete in B. An `insufficient` block
cannot support a MAR branch, while an all-MNAR feature has no complete or
MAR-like condition to support retention.

# On/off rescue and a manual cutoff

This small fixture makes every rule visible. Both conditions have four samples,
and the manual cutoffs are 12.

```{r manual-fixture}
x <- rbind(
    on_off = c(NA, NA, NA, NA, 15, 16, 15, 17),
    mar_both = c(14, 15, NA, 16, 14, NA, 15, 16),
    sparse_mar = c(20, NA, NA, NA, 20, 21, 22, 23),
    sparse_mnar = c(8, NA, NA, NA, 14, 15, 16, 17),
    all_mnar = c(8, NA, NA, NA, 9, NA, NA, NA),
    globally_absent = rep(NA, 8),
    complete_low = c(8, 8, 8, 8, 9, 9, 9, 9)
)
colnames(x) <- paste0("sample_", seq_len(ncol(x)))
condition <- rep(c("A", "B"), each = 4)

fit <- classify_missingness(
    x,
    group = condition,
    cutoffs = c(A = 12, B = 12),
    seed = 1L
)
fit
```

Before classification, a feature missing everywhere is marked `all_missing`
and removed. Every surviving feature fully missing within one condition then
receives one artificial anchor:

1. The minimum finite intensity in each condition is calculated before any
   anchors are inserted.
2. One sample in each fully missing feature-condition block is selected
   uniformly. Selection is with replacement across features.
3. Exactly that cell receives the condition minimum; the others remain `NA`.
4. The feature, condition, selected sample, old and inserted values, and seed
   are recorded.

```{r seed-audit}
fit$seed_log
fit$data["on_off", ]
```

The anchor makes the absent condition classifiable; it is not a completed
imputation. Originally observed cells are preserved exactly. Stable name order
and a local RNG scope make the selected sample reproducible without changing
the caller's global RNG state.

The audit table retains all original feature names, including dropped rows.
The long classification table retains the evidence behind every surviving
feature-condition state.

```{r classification-audit}
fit$feature_status
fit$classifications[
    fit$classifications$feature %in% c("on_off", "sparse_mar", "all_mnar"),
    c("feature", "condition", "observed_count", "missing_count",
      "mean_intensity", "cutoff", "state", "retained", "drop_reason")
]
```

## Result components

| Component | Audit or workflow role |
|---|---|
| `data` | Retained rows in the input representation, with logged rescue anchors inserted. |
| `classifications` | Long feature-condition evidence, state, retention, and drop reason. |
| `groups` | Retained `MNAR`, `MAR`, `complete`, and `MAR_or_complete` feature names for each condition. |
| `feature_status` | One retention decision for every original feature. |
| `cutoffs`, `cutoff_diagnostics` | Condition boundaries, source, method, quality evidence, and warnings. |
| `profiles` | Raw feature statistics, explicit density grid, and smoothing metadata. |
| `seed_log` | Complete provenance for each artificial rescue anchor. |
| `groups_by_sample` | Aligned condition label for every output sample. |
| `call` | Matched classifier call for provenance. |

# The stored missingness profile

For each condition, ImputeFinder estimates missing and complete feature-mean
density curves on one intensity grid. If `n_1 f_1(x)` is the count-weighted
density for blocks with missing values and `n_0 f_0(x)` is the corresponding
density for complete blocks, the displayed profile is

```text
               n_1 f_1(x)
p_missing(x) = -------------------------
               n_1 f_1(x) + n_0 f_0(x)
```

This is the local missing-block proportion represented by a filled stacked
density plot; it is not the average fraction of missing samples. The raw
feature statistics, grid, bandwidth, class counts, warnings, and cutoff
diagnostics are stored in the result. Plotting presents those stored values and
never recovers scientific data from rendered graphics.

```{r stored-evidence}
fit$profiles$A$metadata[c("class_counts", "bandwidth", "kernel")]
head(
    fit$profiles$A$grid[
        , c("intensity", "missing_proportion", "supported")
    ]
)
```

```{r manual-profile, fig.cap="Condition A missingness profile. The vertical line is the recorded manual cutoff; the bottom tick marks the rescued on/off block."}
plot_missingness(fit, "A")
```

This seven-feature fixture tests rule boundaries rather than estimating a
population profile. Its curve is shown to demonstrate stored evidence, the
manual cutoff, and seed provenance; it is too small for scientific breakpoint
interpretation or automatic selection.

# Automatic cutoffs

Manual cutoffs are a permanent reproducibility override and are useful when
domain knowledge or a predeclared analysis fixes the boundary. With
`cutoffs = NULL`, the default, ImputeFinder independently searches each
incomplete condition's profile for the lower, right-hand boundary of its
dominant descending transition. Conditions with no missing observations need
no cutoff. Partial named vectors are also valid: automatic selection fills
only omitted conditions.

The deterministic synthetic matrix below contains two clear but different
transitions. The modular score supplies a fixed mix of incomplete and complete
features, avoiding random vignette output.

```{r automatic-data}
make_condition <- function(n, left, right, score_offset = 0L) {
    index <- seq_len(n)
    means <- qnorm((index - 0.5) / n, mean = 12.5, sd = 2.4)
    score <- (((index + score_offset) * 104729) %% 1000003) / 1000003
    ramp <- pmin(pmax((right - means) / (right - left), 0), 1)
    missing_probability <- 1 - (1 - 0.05) * (1 - 0.68 * ramp)
    missing <- score < missing_probability

    block <- matrix(rep(means, 4L), nrow = n)
    block[cbind(which(missing), rep(4L, sum(missing)))] <- NA_real_
    block
}

feature_count <- 200L
automatic_x <- cbind(
    make_condition(feature_count, left = 10.2, right = 11),
    make_condition(feature_count, left = 13.2, right = 14, score_offset = 37L)
)
rownames(automatic_x) <- sprintf("protein_%04d", seq_len(feature_count))
colnames(automatic_x) <- paste0("automatic_sample_", seq_len(ncol(automatic_x)))
automatic_condition <- rep(c("A", "B"), each = 4)
```

```{r automatic-fit}
automatic_fit <- classify_missingness(
    automatic_x,
    group = automatic_condition
)
round(automatic_fit$cutoffs, 3)
vapply(
    automatic_fit$cutoff_diagnostics,
    function(diagnostic) diagnostic$source,
    character(1)
)
automatic_fit$cutoff_diagnostics$A$method
```

```{r automatic-profile, fig.cap="Automatic condition A profile. The recorded cutoff marks the lower boundary of the dominant descending transition."}
plot_missingness(automatic_fit, "A")
```

An automatic estimate is returned only when total evidence, class evidence,
profile support, a descending trend, and a credible transition all pass the
detector's checks. Flat, one-class, sparse, or ambiguous profiles raise a
condition-specific `imputefinder_cutoff_error` with the profile and diagnostic
attached. Inspect that evidence and supply a named manual cutoff; a failed
automatic call is not permission to invent an endpoint.

# Filtered data and downstream branches

`fit$data` contains only retained features and includes every logged rescue
anchor. `fit$groups` divides those rows independently within each condition
into `MNAR`, `MAR`, `complete`, and the combined `MAR_or_complete` group.

The intended order is:

1. classify the unnormalised log2 matrix;
2. normalise the entire returned `fit$data` object with the method chosen for
   the experiment;
3. split samples by `fit$groups_by_sample`;
4. within each condition, apply an MNAR method only to its `MNAR` rows and a
   MAR method only to its `MAR_or_complete` rows;
5. recombine conditions in original sample order.

The following evaluated code performs the split but deliberately stops before
normalisation or imputation, which are analysis-specific choices and not
ImputeFinder dependencies.

```{r downstream-branches}
split_condition <- function(result, condition_name) {
    samples <- names(result$groups_by_sample)[
        result$groups_by_sample == condition_name
    ]
    condition_data <- result$data[, samples, drop = FALSE]

    list(
        MNAR = condition_data[
            result$groups[[condition_name]]$MNAR,
            ,
            drop = FALSE
        ],
        MAR_or_complete = condition_data[
            result$groups[[condition_name]]$MAR_or_complete,
            ,
            drop = FALSE
        ]
    )
}

downstream <- lapply(names(fit$groups), function(condition_name) {
    split_condition(fit, condition_name)
})
names(downstream) <- names(fit$groups)

vapply(downstream, function(branch) nrow(branch$MNAR), integer(1))
vapply(downstream, function(branch) nrow(branch$MAR_or_complete), integer(1))
c(
    on_off_in_A_MNAR = "on_off" %in% rownames(downstream$A$MNAR),
    on_off_in_B_MAR_or_complete =
        "on_off" %in% rownames(downstream$B$MAR_or_complete)
)
```

Thus `on_off` is routed to an MNAR method in A but to the complete/MAR branch in
B. Complete blocks need no replacement; grouping them with MAR rows supplies a
single non-MNAR branch while preserving their explicit `complete` state in the
classification table.

# `SummarizedExperiment` input

A `SummarizedExperiment` uses the same matrix core. Select the condition column
explicitly and name the assay whenever the object contains more than one. The
returned `data` preserves the representation, unselected assays, row and column
metadata, and sample order while filtering rows and replacing only the selected
assay with the seed-modified values.

```{r summarized-experiment}
se <- SummarizedExperiment::SummarizedExperiment(
    assays = list(log2_intensity = x),
    colData = data.frame(
        condition = condition,
        row.names = colnames(x)
    )
)
fit_se <- classify_missingness(
    se,
    group_col = "condition",
    assay = "log2_intensity",
    cutoffs = c(A = 12, B = 12)
)
identical(
    SummarizedExperiment::assay(fit_se$data, "log2_intensity"),
    fit$data
)
```

# Limitations

- The cutoff is a model assumption. Biology, acquisition, and preprocessing can
  produce patterns without one descending intensity transition.
- Automatic breakpoint selection needs both incomplete and complete feature
  evidence. Structured failure is an expected result for unidentifiable data.
- Arithmetic means are sensitive to strong observations by design.
- Rescue anchors are artificial values. Preserve `seed_log` with the analysis
  and assess their influence downstream.
- Classification is heuristic. ImputeFinder does not prove mechanisms, choose
  an imputation algorithm, or validate the final biological analysis.

The deterministic repository validation covers intensity-independent MAR,
intensity-dependent MNAR, condition-specific on/off features, several group
sizes and cutoffs, rescue seeds, and name permutations. Synthetic evidence
supports those tested behaviors but does not establish performance for every
proteomics experiment.

# Related methods and references {.unnumbered}

- Karpievitch YV, Dabney AR, Smith RD (2012). [Normalization and missing value
  imputation for label-free LC-MS
  analysis](https://doi.org/10.1186/1471-2105-13-S16-S5). *BMC Bioinformatics*
  13(Suppl 16):S5.
- Lazar C, Gatto L, Ferro M, Bruley C, Burger T (2016). [Accounting for the
  multiple natures of missing values in label-free quantitative proteomics data
  sets to compare imputation
  strategies](https://doi.org/10.1021/acs.jproteome.5b00981). *Journal of
  Proteome Research* 15:1116-1125.

# Session information {.unnumbered}

```{r session-info}
suppressWarnings(sessionInfo())
```
