---
title: "Introduction to imputefinder"
author: "Emir Turkes"
date: "`r Sys.Date()`"
package: imputefinder
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to imputefinder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`imputefinder` provides a flexible method for distinguishing between
**Missing Not At Random (MNAR)** and **Missing At Random (MAR)** values
in quantitative proteomics data such as label-free quantification (LFQ).
For each experimental condition the package inspects the empirical
distribution of feature mean intensities and uses elbow detection on
the cumulative intensity curve to derive an intensity cutoff. Features
with mean intensity below the cutoff are classified as MNAR; remaining
features with enough non-missing values are classified as MAR.

The package exposes a single user-facing entry point,
`classify_missingness()`, which operates on a
`SummarizedExperiment` whose `colData` contains a `condition`
column.

# Using `classify_missingness()`

```{r load}
library(imputefinder)
library(SummarizedExperiment)
```

`classify_missingness()` enforces two preconditions on its input. First,
the supplied object must inherit from `SummarizedExperiment`; any other
object is rejected with an informative error.

```{r non-se-input}
res <- try(classify_missingness(matrix(1:6, nrow = 2)), silent = TRUE)
inherits(res, "try-error")
```

Second, the assay must actually contain missing values, since the
classification is only meaningful in that case.

```{r no-missing}
counts <- matrix(
  seq_len(12),
  nrow = 4,
  dimnames = list(
    paste0("p", seq_len(4)),
    paste0("s", seq_len(3))
  )
)
se <- SummarizedExperiment(
  assays = list(intensity = counts),
  colData = DataFrame(condition = c("A", "A", "B"))
)
res <- try(classify_missingness(se), silent = TRUE)
inherits(res, "try-error")
```

In a typical analysis you would pass a `SummarizedExperiment` of
log-intensities with a `condition` column in `colData()` and several
`NA` values in the assay; the call returns a named list with the
filtered `SummarizedExperiment`, the per-condition cutoffs, and the
sets of features classified as MAR and MNAR. The `return_plot = TRUE`
argument additionally yields a list of `ggplot` objects showing the
cumulative intensity curve and the detected elbow per condition.

# Session information

```{r sessioninfo}
sessionInfo()
```
